# Introduction to Linear Programming

```{r, results = 'asis', echo = FALSE}
source("R/linear_program.R", local = knitr::knit_global())
source("R/linear_program_general.R", local = knitr::knit_global())
```
<script src="https://www.geogebra.org/apps/deployggb.js"></script>


Linear programming is a technique for the optimization of a linear objective function, subject to linear equality and linear inequality constraints. A **linear program** or a **linear programming problem** is a problem of the following form:

\begin{equation}
  \begin{aligned}
    \mbox{optimize: } && c_1 x_1 + \dots + c_n x_n & \\
    \mbox{subject to: } 
      && a_{11} x_1 + \dots + a_{1n} x_n & \lesseqgtr b_1 \\
      && a_{21} x_1 + \dots + a_{2n} x_n & \lesseqgtr b_2 \\
      && \vdots &  \\
      && a_{m1} x_1 + \dots + a_{mn} x_n & \lesseqgtr b_m,
  \end{aligned}
  (\#eq:intro-lp)
\end{equation}
where the symbol $\lesseqgtr$ stands for $\leq$ or $=$ or $\geq$, $a_{ij}, b_i, c_j$ are real numbers, and $x_j$ are variables.

The variables $x_1, \dots, x_n$ are called **decision variables**. The linear combination $\zeta := c_1 x_1 + \dots + c_n x_n$ is called the **objective function**. Each of the equalities/inequalities $a_{i1} x_1 + \dots + a_{in} x_n \lesseqgtr b_i$ is called a **(linear) constraint**. Our goal is to either *minimize* or *maximize* the objective function subject to the constraints.

A proposal of specific values for the decision variables is called a **solution**. A solution is said to be **feasible** if it satisfies all the constraints. A feasible solution is said to be **optimal** if $\zeta$ attains the optimal value at it. Thus, to solve a linear program means to *find an optimal solution* to the problem. If a problem has no feasible solutions, then the problem is called **infeasible**. If a problem has feasible solutions with arbitrarily large or arbitrary small objective values then the problem is called **unbounded**.

:::{.remark}

We do not allow strict inequalities $<$ or $>$ in a linear program as linear functions do not always achieve maxima/minima on open sets. Consider the following simple example.

```{r, results = 'asis', echo = FALSE}
ob <- new_linear_combination(10)
c1 <- new_constraint(ob, "<", 20)
c2 <- new_constraint(new_linear_combination(5), "\\geq", 0)
cat(str_math(new_linear_program_general(ob, list(c1, c2))))
```

On the feasible set $[0, 2)$, the function $\zeta(x_1) = x_1$ never attains absolute maxima.
Changing the inequality $<$ to $\leq$ gives us an optimal feasible solution $x_1 = 2$. When all the inequalities are either $\le$, $=$, or $\ge$, the set of feasible solutions is closed. If, in addition, the LP is bounded then the set of feasible solutions is compact. On compact sets a continuous (in particular, linear) function, always attains a maxima and a minima by a generalization of the [extreme value theorem](https://en.wikipedia.org/wiki/Extreme_value_theorem) for [higher dimensions](https://math.stackexchange.com/a/881569).

:::

## Resource allocation problem
The following is an example of a *resource allocation problem*, a very common application of linear programming. 
The decision variables $x_j$ denote the amount of a certain resource/product $j$, the coefficients $c_j$ are the profit per unit quantity of $j$, and the constraints are certain upper bounds on the production of the resource/quantity.

::: {.example #resource-allocation-problem}
You run a company that makes two products (say, $P_1$ and $P_2$) using two machines (say, $M_1$ and $M_2$). Each unit of $P_1$ that is produced requires 80 minutes processing time on machine $M_1$ and 25 minutes processing time on machine $M_2$. Each unit of $P_2$ that is produced requires 20 minutes processing time on machine $M_1$ and 75 minutes processing time on machine $M_2$. Both machines are available for a maximum of 200 minutes every day. The profit per unit of $P_1$ is 25 units and the profit per unit of $P_2$ is 30 units. Company policy is to determine the production quantity of each product in such a way as to maximize the total profit given that the available resources should not be exceeded.^[Problem from https://ocw.mit.edu/courses/sloan-school-of-management/15-053-optimization-methods-in-management-science-spring-2013/tutorials/MIT15_053S13_tut01.pdf]

```{r, echo=FALSE, comment=NA}

df <- data.frame(
  P1 = c(80, 20),
  P2 = c(25, 75),
  Upper_Bounds = c(200, 200)
)
rownames(df) <- c("M1", "M2")
df
```
:::

We can formulate the above problem as the following linear program.

```{r, results = 'asis', echo = FALSE}
ex_1_1_lp <- new_linear_program(
  obj = c(25, 30),
  A = matrix(c(80, 20, 25, 75), nrow = 2, byrow = TRUE),
  b = c(200, 200),
  variables = c('x', 'y')
)

cat(str_math(ex_1_1_lp))
```

Here $x$, $y$ are the units of $P_1$ and $P_2$ produced each day. One way to attempt this problem is to assume that we'll use both the machines as much as possible i.e. that the two inequalities are in fact equalities. We can see then that the common solution to the two is $(x_1, x_2) = (2, 2)$. For this solution the total profit is $110$ units.

```{r echo=FALSE}

library(ggplot2)

ggplot() + 
  labs(title = "Resource Allocation Problem Constraints") +
  xlim(0, 200/25) +
  ylim(0, 200/20) +
  geom_vline(xintercept=c(0)) +
  geom_hline(yintercept=c(0)) +
  theme_bw() + 
  geom_polygon(
      data = data.frame(
        x = c(0, 0, 200/80, Inf),
        y = c(0, 200/20, 0, 0)),
      aes(x = x, y = y, fill = "Constraint 1: 80x + 20y <= 200"),
      inherit.aes = FALSE, alpha = 0.4) + 
  geom_polygon(
      data = data.frame(
        x = c(0, 0, 200/25, Inf),
        y = c(0, 200/75, 0, 0)),
      aes(x = x, y = y, fill = "Constraint 2: 25x + 75y <= 200"),
      inherit.aes = TRUE, alpha = 0.4) +
  scale_fill_discrete(name = "Constraints") +
  geom_point(aes(x=2, y=2), color = "black") +
  geom_text(aes(x=2, y=2), label="(2, 2)", vjust=-.5, hjust=-.5)
```

::: {.exercise}
How to we know that this is the maximum profit possible? Is it possible to increase the profit by not using either of the two machines at their full capacity?
:::


::: {.exercise}
Consider the setup in Example \@ref(resource-allocation-problem) again. Suppose in addition, you can only store a total of 3 units of $P_1$ and $P_2$ combined each day so your production quantity should not exceed that amount. What is the new LP and the new optimal solution?
Note that the solution $(x, y) = (2, 2)$ is no longer possible as $x + y > 3$. So, we cannot use both the machines to their maximum capacity.
:::



