# (PART) Part I - Linear Programming {-}

# Introduction to Linear Programming

```{r, results = 'asis', echo = FALSE}
source("R/linear_program_general.R", local = knitr::knit_global())
```


Linear programming is a technique for the optimization of a linear objective function, subject to linear equality and linear inequality constraints. A *linear program* or a *linear programming problem* is a problem of the following form:

\begin{equation}
  \begin{aligned}
    \mbox{optimize: } && c_1 x_1 + \dots + c_n x_n & \\
    \mbox{subject to: } 
      && a_{11} x_1 + \dots + a_{1n} x_n & \lesseqgtr b_1 \\
      && a_{21} x_1 + \dots + a_{2n} x_n & \lesseqgtr b_2 \\
      && \vdots &  \\
      && a_{m1} x_1 + \dots + a_{mn} x_n & \lesseqgtr b_m,
  \end{aligned}
  (\#eq:intro-lp)
\end{equation}
where the symbol $\lesseqgtr$ stands for $\leq$ or $=$ or $\geq$, $a_{ij}, b_i, c_j$ are real numbers, and $x_j$ are variables.

The variables $x_1, \dots, x_n$ are called *decision variables*. The linear combination $\zeta := c_1 x_1 + \dots + c_n x_n$ is called the *objective function*. Each of the equalities/inequalities $a_{i1} x_1 + \dots + a_{in} x_n \lesseqgtr b_i$ is called a *(linear) constraint*. Our goal is to either *minimize* or *maximize* the objective function subject to the constraints.

A proposal of specific values for the decision variables is called a *solution*. A solution is said to be *feasible* if it satisfies all the constraints. A feasible solution is said to be *optimal* if $\zeta$ attains the optimal value at the solution. Thus, to solve a linear program means to *find an optimal solution* to the problem. If a problem has no feasible solutions, then the problem is called *infeasible*. If a problem has feasible solutions with arbitrarily large or arbitrary small objective values then the problem is called *unbounded*.

:::{.remark}

We do not allow strict inequalities $<$ or $>$ in a linear program as linear functions do not always achieve maxima/minima on open sets. Consider the following simple example.

```{r, results = 'asis', echo = FALSE}
ob <- new_linear_combination(10)
c1 <- new_constraint(ob, "<", 20)
c2 <- new_constraint(new_linear_combination(5), "\\geq", 0)
cat(str_math(new_linear_program_general(ob, list(c1, c2))))
```

On the feasible set $[0, 2)$, the function $\zeta(x_1) = x_1$ never attains absolute maxima.
Changing the inequality $<$ to $\leq$ gives us an optimal feasible solution $x_1 = 2$. When all the inequalities are either $\le$, $=$, or $\ge$, the set of feasible solutions is closed. If, in addition, the LP is bounded then the set of feasible solutions is compact. On compact sets a continuous (in particular, linear) function, always attains a maxima and a minima by a generalization of the [extreme value theorem](https://en.wikipedia.org/wiki/Extreme_value_theorem) for [higher dimensions](https://math.stackexchange.com/a/881569).

:::

## Resource allocation problem
It is useful to think of the objective function as the *profit* function, the decision variable $x_j$ as the amount of a physical quantity of type $j$, the coefficients $c_j$ as being the profit made from selling a unit quantity of type $j$, and each constraint as describing an upper bound on the production of an intermediate quantity. This is called a *resource allocation problem*.

::: {.example #dunder-mifflin-lp-intro}
Consider the following situation. Suppose you are the reginal manager of Dunder Mifflin Paper Company and you want to expand your branch. Given your office space restrictions, you can hire at most 5 new staff to your office. You can either hire some salespersons with a salary of 65K per month or trainees with a salary of 15K per month. You predict that your profits will go up by 45K for each salesperson and by 5k for each trainee. How many salespersons and trainees should you hire to maximize your predicted revenue if you can pay out total salaries of maximum 175K per month? 

:::