# The Simplex Method - Example

The simplex method jumps from vertex of the feasible region to find the optimal solution. It is best demonstrated using an example. Consider the following linear program.

\begin{equation}
```{r, simplex-example, results = 'asis'}
simplex_method_example <- new_linear_program(
  obj = c(1, 1),
  A = matrix(c(1, 2, 4, 3), nrow = 2, byrow = TRUE),
  b = c(10, 20),
  variables = c('x_1', 'x_2')
)

cat(str(simplex_method_example))
```
(\#eq:simplex-example)
\end{equation}

The optimal solution is $(x_1, x_2) = (2, 4)$ with objective value $6$. There are 6 different points at which the lines corresponding to the constraints intersect. We'll start with the vertex $(0,0)$ and make our way up to $(2, 4)$.

```{r, simplex-example-plot}
library(ggplot2)

simplex_example_plot <- ggplot() + 
  xlim(0, max(10, 20/4)) +
  ylim(0, max(10/2, 20/3)) +
  geom_vline(xintercept=c(0), alpha = 0.4) +
  geom_hline(yintercept=c(0), alpha = 0.4) +
  theme_bw() + 
  geom_polygon(
      data = data.frame(
        x = c(0, 0, 10/1),
        y = c(0, 10/2, 0)),
      aes(x = x, y = y, fill = "1: x + 2y <= 10"),
      inherit.aes = FALSE, alpha = 0.4) + 
  geom_polygon(
      data = data.frame(
        x = c(0, 0, 20/4),
        y = c(0, 20/3, 0)),
      aes(x = x, y = y, fill = "2: 4x + 3y <= 20"),
      inherit.aes = TRUE, alpha = 0.4) +
  scale_fill_brewer(name = "Constraints", palette = "Pastel2")

simplex_example_plot +
  geom_point(aes(x=2, y=4)) +
  geom_point(aes(x=10/1, y=0)) +
  geom_point(aes(x=0, y=10/2)) +
  geom_point(aes(x=20/4, y=0)) +
  geom_point(aes(x=0, y=20/3)) +
  geom_point(aes(x=0, y=0))
```

The first, and most non-trivial, algebraic trick is to introduce *slack variables*  $x_3$, $x_4$ using the constraints as follows and create a new linear program called a *dictionary* which is equivalent to \@ref(eq:simplex-example)

\begin{equation}
``` {r, resource-example-dictionary, results="asis"}
x <- new_dictionary(
  objective = c(0, 1, 1),
  coefficients = matrix(c(1, 2, 4, 3), nrow = 2, byrow = TRUE),
  constants = c(10, 20),
  non_basic_variables = c("x_1", "x_2"),
  basic_variables = c("x_3", "x_4")
)
cat(str(x))
```
(\#eq:simplex-example-dictionary)
\end{equation}

Using these new variables, the points of intersection be written as intersections of the lines $x_i = 0$ and $x_j = 0$, as $i$ and $j$ vary over the index set $\{1, 2, 3, 4\}$. 

```{r, simplex-example-intersections-labeled}
library(ggplot2)

simplex_example_plot +
  geom_point(aes(x=0, y=0, color = "x_1, x_2")) +
  geom_point(aes(x=0, y=10/2, color = "x_1, x_3")) +
  geom_point(aes(x=0, y=20/3, color = "x_1, x_4")) +
  geom_point(aes(x=10/1, y=0, color = "x_2, x_3")) +
  geom_point(aes(x=20/4, y=0, color = "x_2, x_4")) +
  geom_point(aes(x=2, y=4, color = "x_3, x_4")) +
  scale_colour_brewer("Intersection points", palette = "Set1")
```

1. We start with the guess: $x_1 = 0, x_2 = 0, x_3 = 10, x_4 = 20$. 
    1. At this point, $\zeta = 0$. We can increase $\zeta$ by increasing either $x_1$ or $x_2$. Let's choose $x_1$ and leave $x_2$ fixed at $0$.
    2. To not violate the positivity of $x_3$, $x_4$, we can only increase $x_1$ enough that both $10 \ge x_1$ and $20 \ge 4x_1$ remain true. The maximum such $x_1$ is $5$ and this makes $x_4 = 0$.
    3. We rewrite $x_1$, $x_3$, and $\zeta$ in terms of the new variables that are $0$, namely, $x_2$ and $x_4$ to create a new dictionary:
    \begin{equation}
    ``` {r, simplex-example-step-1, results="asis"}
    x <- new_dictionary(
      objective = c(5, -1/4, 1/4),
      coefficients = matrix(c(-1/4, 5/4, 1/4, 3/4), nrow = 2, byrow = TRUE),
      constants = c(5, 5),
      basic_variables = c("x_3", "x_1"),
      non_basic_variables = c("x_4", "x_2")
    )
    cat(str(x))
    ```
    (\#eq:simplex-example-dictionary-step-1)
    \end{equation}
2. Our new guess is: $x_1 = 5, x_2 = 0, x_3 = 5, x_4 = 0$. 
    1. The coefficient of $x_4$ in the objective function is negative. Increasing it will decrease the objective value. The only variable we can increase in the objective function is $x_2$.
    2. To not violate the positivity of $x_1$, $x_3$, we can only increase $x_3$ enough that both $5 \ge 1.25x_2$ and $5 \ge 0.75x_2$ remain true. The maximum such $x_2$ is $\min(5/1.25, 5/0.75) = 5/1.25 = 4$ and this makes $x_3 = 0$.
    3. We rewrite $x_2$, $x_1$, and $\zeta$ in terms of the new variables that are $0$, namely, $x_3$ and $x_4$ to create a new dictionary:
    \begin{equation}
    ``` {r, simplex-example-step-2, results="asis"}
    x <- new_dictionary(
      objective = c(6, -1/5, -1/5),
      coefficients = matrix(c(-1/5, 4/5, 2/5, -3/5), nrow = 2, byrow = TRUE),
      constants = c(4, 2),
      basic_variables = c("x_2", "x_1"),
      non_basic_variables = c("x_4", "x_3")
    )
    cat(str(x))
    ```
    (\#eq:simplex-example-dictionary-step-2)
    \end{equation}
3. Our new guess is: $x_1 = 2, x_2 = 4, x_3 = 0, x_4 = 0$. 
    1. The coefficients of both the variables that appear in $\zeta$ are negative. It is not possible to increase either variable without decreasing the objective value. The simplex algorithm halts. 

The optimal objective value $6$ is the constant coefficient in the final $\zeta$ and it is attained at $(x_1, x_2) = (2, 4)$.


```{r, simplex-example-intersections-traced}
library(ggplot2)

simplex_example_plot +
  geom_point(aes(x=0, y=0, color = "Guess 1: x_1, x_2")) +
  geom_point(aes(x=20/4, y=0, color = "Guess 2: x_2, x_4")) +
  geom_point(aes(x=2, y=4, color = "Guess 3: x_3, x_4")) +
  scale_colour_brewer("Guesses", palette = "Set1")
```


## Summary

Here's a summary of the simplex method used in the above example, with a few technical terms introduced.

1. *Setup:*
    1. Introduce slack variables.
    2. Create dictionary.
    3. Set all the decision variables to 0. 
    4. Calculate the values of the slack variables. This is our initial guess.
    5. Call the decision variables the non-basic variables and the slack variables the basic variables.
2. *Choose an entering variable:*
    1. Choose a non-basic variable whose coefficient in the objective function is positive.
      - **If no such variable exists, the algorithm halts and the current guess is the optimal solution.**
    2. Call this variable the entering variable.
    3. Add the entering variable to the list of basic variables and remove it from the list of non-basic variables.
3. *Find the leaving variable:* 
    1. Find the maximum value the entering variable can be increased to without violating any of the positivity constraints.
    2. This maximum value will force the corresponding basic variable to become zero. Call this the leaving variable.
    3. Add the leaving variable to the list of non-basic variables and remove it from the list of basic variables.      
4. *Update dictionary:*
    2. Rewrite the entering variable in terms of the leaving variable.
    3. Rewrite the objective function and all the other basic variables in terms of the new non-basic variables.
5. *Recalculate guess:*
    1. Set the maximum value found above as the new value of the entering variable.
    2. Set the leaving variable to 0.
    1. Recalculate values of the other basic variables and the objective function.
    2. Update the guess.
5. *Go back to Step 2*.